Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
added details to complexity section

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T = 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
Ignoring the possibility of coalescent events involving more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$,
which is also equal to the mean number of ancestral recombination events that occur on a branch of the marginal tree,
since we have taken one crossover per individual per generation.
Not all such recombinations actually change the tree topology,
so four times this gives us an upper bound on the expected number of additional edges.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
With $T=20N$, this reduces the initial storage of $120 N^2$ items to
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
$6N + 32 N \log(2N)$;
at $N=10^4$, a factor of 3,700.

@@

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$,
there would also be around $\mu 2NT$ mutations (and the same number of sites).
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations,
and the same savings as we obtained from edges above,
even without considering the
computational burden of propagating mutations forwards across generations.

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
so $M$ will typically be $O(N \log N)$ as well.

%% simplify run time
How does the computation \emph{time} required for simplification scale?
Simply because it must process each edge,
% below per discussion
% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
the simplification algorithm is at least linear in the number of edges of the input tree sequence.
Empirically the algorithm is exactly linear,
as seen in Figure~\ref{fig:simplify_complexity}B,
where we show shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C,

>>>>

%% Memory reduction thanks to simplification
Consider a simulation of a Wright-Fisher population of $2N$ haploid individuals
using Algorithm~\algref{W} for $T$ generations,
with simplify interval $s = 1$ so that redundant information is removed after every generation.
Since every chromosome inherits material from both parents,
without simplification this would produce tables of
$2NT$ nodes and
$4NT$ edges.
Suppose that $T$ is large enough that all samples coalesce within the simulation with high probability,
($T = 20N$, say).
After simplification, we are left with the tree sequence describing the history
of only the current generation of $2N$ individuals.
If no coalescent events involve more than two lineages, this tree sequence has $4N-2$ edges to describe the leftmost tree;
and each time the marginal tree changes along the sequence,
four edges end and four new edges begin
(except for changes affecting the root, which require fewer; see \citet{kelleher2016efficient}).
Coalescent theory tells us that
the expected total length of the edges of a marginal tree is approximately $4N\log(2N)$.
Since we have taken one crossover per individual per generation (as is typical for a single chromosome),
and we are measuring tree lengths in units of generations,
the mean number of ancestral recombination events that occur somewhere on the marginal trees
per unit of genetic length.
Any such ancestral recombination might change the marginal gene tree;
if it does, it will add four new edges.
Not all such recombinations actually change the tree topology,
so at most we expect $16N\log(2N)$ new edges after those required to describe the first tree.
Similarly, not every new edge derives from a never-before-seen node,
but the number of nodes is at most equal to the number of edges plus the sample size.
This reduces the initial storage of $120 N^2$ items to
% 4N - 2 + 2N + 4N log(2N) * 4 * 2 = 6N + 32 N log(2N)
the space required after simplification of roughly
\begin{align}
    6N + 32 N \log(2N) .
\end{align}
At $N=10^4$, this reduces the space required to store the history of the entire population
by a factor of 3,700.

@@

%% Above was edges; same holds for memory usage savings due to mutations
If we were to add mutations in forwards time
under the infinite-sites model with total mutation rate per generation $\mu$ (i.e., per meiosis),
there would also be around $\mu 2NT$ mutations, and the same number of sites,
simply because there was a total of $2NT$ meioses.
Since mutations fall as recombinations do on the marginal trees,
the number of mutations is proportional to the total length of all the marginal trees,
and so by the same argument as for the number of edges,
adding them after simplification results in only around $\mu 4 N \log(2N)$ mutations.
With $N=10^4$ and $T=20N$ as above,
this implies that adding neutral mutations afterwards reduces the number of mutations that must be generated
by a factor of 10,000.
This could result in substantial time savings,
even without considering the
computational burden of propagating mutations forwards across generations.

Since each mutation is stored only as a single row in the mutation table,
and at most one row in the site table, the space required for $M$ mutations is $O(M)$;
combined with the $O(N \log N)$ storage for edges and nodes of a simplified tree sequence above,
this implies that the full results of a simulation of $N$ individuals
having $M$ mutations can be stored in $O(N \log N + M)$ space.
In simulations of whole chromosomes, the number of mutations per chromosome per generation is $O(1)$,
so $M$ will typically be $O(N \log N)$ as well.

%% simplify run time
How does the computation \emph{time} required for simplification scale?
Simply because it must process each edge,
% below per discussion
% https://github.com/petrelharp/ftprime_ms/pull/37#discussion_r156617283
the simplification algorithm is at least linear in the number of edges of the input tree sequence.
Empirically the algorithm is exactly linear,
as seen in Figure~\ref{fig:simplify_complexity}B,
which shows the time required to simplify increasingly large subsets of a large tree sequence.
When simplifying the result of a forwards-time sequence, the number of edges is the main contributing factor.
Suppose on the other hand we want to
simplify an already-minimal but large tree sequence with $N$ nodes
to a subsample of size $n$.
How does the required time scale with $n$?
In this case, the computation is dominated by the size of the output tree sequence,
which grows with $\log(n)$, as shown in Figure~\ref{fig:simplify_complexity}C,
